Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-09-03 manuscript.qmd
make vars in data section dynamic

from scripts import shared, paths, COL, get_lnl_cols

@@

1. In section @sec-data we present a multi-centric dataset on lymph node involvement in 833 OPSCC patients. Based on the dataset the main risk factors for contralateral lymph node involvement are identified and the requirements for the bilateral model extension are described (@sec-requirements).

@@

To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of 833<!-- num patients --> patients with newly diagnosed oropharyngeal squamous cell carcinomas [@ludwig_dataset_2022;@ludwig_multi-centric_2023]. It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subside, and lateralization of the primary tumor, and HPV p16 status. Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview. The data from the Inselspital Bern and the Centre Léon Bérard only consist of patients who received a neck dissection. Since this treatment is more commonly chosen for early T-category patients, they also make up a larger portion of the respective dataset. The majority of patients from the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital was treated with definitive radiotherapy. For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.

@@

The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated. For example, midline extension was present in 48% of advanced T-category tumors but only in 7% of early T-category tumors. The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors. However, the bottom right panel of @fig-data-strat indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors. The figure considers only patients with lateralized tumors not extending over the midline. For early T-category and healthy ipsilateral levels I-V, only 1% (1 out of 84 patients) shows involvement of contralateral level II. This increases to 5% (9 out of 165 patients) if ipsilateral level II is involved, to 15% (9 out of 60 patients) if ipsilateral levels II and III are involved, and to 23% (6 out of 26 patients) for advanced T-category with ipsilateral levels II and III involved.<!-- TODO: make values here dynamic. -->

@@

grouped = contra_inv.groupby(by=["t_stage", "ipsi", "midext"])
num_involved = grouped.sum()
total = grouped.count()
percent_involved = num_involved / total
idx = total.index.rename(["T-cat.", "ipsi", "Mid. ext."])

@@

involved = num_involved.join(
  100 * percent_involved,
  rsuffix=" (%)",
).sort_index(axis="columns")
involved.columns = pd.MultiIndex.from_product(
  [["I", "II", "III", "IV"],
   ["n", "%"]],
  names=["LNL", ""]
)
involved["total", "n"] = total["I"]
(
  involved
  .reset_index()
  .sort_values("ipsi", ascending=True)
  .sort_values("T-cat.", ascending=False)
  .style
  .format(precision=2)
  .apply(right_align, subset=involved.columns[0:])
  .apply(highlight_t_stage, subset=["T-cat."])
  .apply(highlight_bool, subset=["Mid. ext."])
  .apply(highlight, subset=["ipsi"], mapping={
    "0": COLORS["green"],
    "1": COLORS["orange"],
    "≥ 2": COLORS["red"],
  })
  .hide()
)
```

>>>>

from scripts import shared, paths
from scripts.shared import COL, get_lnl_cols

@@

1. In section @sec-data we present a multi-centric dataset on lymph node involvement in {{< var data.num_patients >}} OPSCC patients. Based on the dataset the main risk factors for contralateral lymph node involvement are identified and the requirements for the bilateral model extension are described (@sec-requirements).

@@

To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas [@ludwig_dataset_2022;@ludwig_multi-centric_2023]. It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subside, and lateralization of the primary tumor, and HPV p16 status. Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview. The data from the Inselspital Bern and the Centre Léon Bérard only consist of patients who received a neck dissection. Since this treatment is more commonly chosen for early T-category patients, they also make up a larger portion of the respective dataset. The majority of patients from the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital was treated with definitive radiotherapy. For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.

@@

The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated. For example, midline extension was present in {{< var data.advanced_with_midext_percent >}}% of advanced T-category tumors but only in {{< var data.early_with_midext_percent >}}% of early T-category tumors. The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors. However, the bottom right panel of @fig-data-strat indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors. The figure considers only patients with lateralized tumors not extending over the midline. For early T-category and healthy ipsilateral levels I-V, only {{< var data.early_ipsin0_cII_percent >}}% ({{< var data.early_ipsin0_cII_match >}} out of {{< var data.early_ipsin0_cII_total >}} patients) shows involvement of contralateral level II. This increases to {{< var data.early_ipsiII_cII_percent >}}% ({{< var data.early_ipsiII_cII_match >}} out of {{< var data.early_ipsiII_cII_total >}} patients) if ipsilateral level II is involved, to {{< var data.early_ipsiIIandIII_cII_percent >}}% ({{< var data.early_ipsiIIandIII_cII_match >}} out of {{< var data.early_ipsiIIandIII_cII_total >}} patients) if ipsilateral levels II and III are involved, and to {{< var data.late_ipsiIIandIII_cII_percent >}}% ({{< var data.late_ipsiIIandIII_cII_match >}} out of {{< var data.late_ipsiIIandIII_cII_total >}} patients) for advanced T-category with ipsilateral levels II and III involved.

@@

grouped = contra_inv.groupby(by=["t_stage", "ipsi", "midext"], dropna=False)
num_involved = grouped.sum()
total = grouped.count()
percent_involved = num_involved / total
idx = total.index.rename(["T-cat.", "ipsi", "Mid. ext."])

@@

involved = num_involved.join(
  100 * percent_involved,
  rsuffix=" (%)",
)
involved.columns = pd.MultiIndex.from_product(
  [["I", "II", "III", "IV"],
   ["n", "%"]],
  names=["LNL", ""]
)
involved["total", "n"] = total["I"]
(
  involved
  .reset_index()
  .sort_values(
    by=["T-cat.", "ipsi", "Mid. ext."],
    ascending=[False, True, True],
  )
  .style
  .format(precision=2)
  .apply(right_align, subset=involved.columns[0:])
  .apply(highlight_t_stage, subset=["T-cat."])
  .apply(highlight_bool, subset=["Mid. ext."])
  .apply(highlight, subset=["ipsi"], mapping={
    "0": COLORS["green"],
    "1": COLORS["orange"],
    "≥ 2": COLORS["red"],
  })
  .hide()
)
```
